The Strucplot Framework:\\ Visualizing Multi-way Contingency Tables with \pkg{vcd}

### 2005-11-02 trunk/vcd/paper/JSS.Rnw
Final touches (table 2 is now sorted). Submission release!
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@491 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

The strucplot framework is highly modularized: Figure \ref{fig:struc}
shows the hierarchical relationship between the various components. 
On the lowest level, there are several groups of workhorse and 
parameter functions that directly or indirectly influence the final
appearance of the plot. These are examples of `graphical appearance control'
(`grapcon') functions. They are created by generating functions
(`grapcon generators'), allowing
flexible parameterization and extensibility (Figure \ref{fig:struc}
only shows the generators). The generator names
follow the naming convention \code{\emph{group\_foo}()},  
where (\code{\emph{group}}) reflects the group the
generators belong to (strucplot core, labeling,
legend, shading, or spacing). The workhorse functions (created by 
\code{struc\_\emph{foo}()}, 
\code{labeling\_\emph{foo}()}, and \code{legend\_\emph{foo}()}) 
directly produce graphical output (i.e., ``add ink to the canvas''), whereas
the parameter functions
(created by \code{spacing\_\emph{foo}()} and \code{shading\_\emph{foo}()}) compute
graphical parameters used by the others. The grapcon functions returned by 
\code{struc\_\emph{foo}()} implement the core functionality, creating the tiles and their
content.  On the second level of the framework, a suitable combination
of the low-level grapcon functions (or, alternatively, corresponding generating functions)
is passed as ``hyperparameters'' to \codefun{strucplot}.
This central function 
sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
and coordinates the specified core, labeling, shading, and spacing functions to produce 
the plot. On the third level, we provide
several convenience functions such as \codefun{mosaic},
\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
interface \codefun{strucplot} through sensible parameter defaults
and support for model formulae. Finally, on the fourth
level, there are `related' \pkg{vcd} functions (such as \codefun{cotabplot}
and the \codefun{pairs} methods for table objects) 
arranging collections of plots of the strucplot 
framework into more complex displays (e.g., by means of panel functions).

>>>>

The strucplot framework is highly modularized: Figure \ref{fig:struc}
shows the hierarchical relationship between the various components. 
On the lowest level, there are several groups of workhorse and 
parameter functions that directly or indirectly influence the final
appearance of the plot (see Table \ref{tab:grapcons} for an overview). 
These are examples of `\underline{gr}aphical \underline{ap}pearance \underline{con}trol'
(`grapcon') functions. They are created by generating functions
(`grapcon generators'), allowing
flexible parameterization and extensibility (Figure \ref{fig:struc}
only shows the generators). The generator names
follow the naming convention \code{\emph{group\_foo}()},  
where (\code{\emph{group}}) reflects the group the
generators belong to (strucplot core, labeling,
legend, shading, or spacing). The workhorse functions (created by 
\code{struc\_\emph{foo}()}, 
\code{labeling\_\emph{foo}()}, and \code{legend\_\emph{foo}()}) 
directly produce graphical output (i.e., ``add ink to the canvas''), whereas
the parameter functions
(created by \code{spacing\_\emph{foo}()} and \code{shading\_\emph{foo}()}) compute
graphical parameters used by the others. The grapcon functions returned by 
\code{struc\_\emph{foo}()} implement the core functionality, creating the tiles and their
content.  On the second level of the framework, a suitable combination
of the low-level grapcon functions (or, alternatively, corresponding generating functions)
is passed as ``hyperparameters'' to \codefun{strucplot}.
This central function 
sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
and coordinates the specified core, labeling, shading, and spacing functions to produce 
the plot. On the third level, we provide
several convenience functions such as \codefun{mosaic},
\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
interface \codefun{strucplot} through sensible parameter defaults
and support for model formulae. Finally, on the fourth
level, there are `related' \pkg{vcd} functions (such as \codefun{cotabplot}
and the \codefun{pairs} methods for table objects) 
arranging collections of plots of the strucplot 
framework into more complex displays (e.g., by means of panel functions).

\begin{table}
  \begin{tabular}{|l|l|l|}
    \hline
    \textbf{Group} & \textbf{Grapcon generator} & \textbf{Description}\\\hline
    strucplot & \codefun{struc\_assoc} & core function for association plots\\
    core      & \codefun{struc\_mosaic} & core function for mosaic plots\\
              & \codefun{struc\_sieve} & core function for sieve plots\\\hline\hline
     labeling & \codefun{labeling\_border} & border labels\\
              & \codefun{labeling\_cboxed} & centered labels with
              boxes, all labels clipped,\\
              && and on top and left border\\
              & \codefun{labeling\_cells} & cell labels\\
              & \codefun{labeling\_conditional} & border labels 
                                                  for conditioning variables\\
              && and cell labels for conditioned variables.\\
              & \codefun{labeling\_doubledecker} & draws labels for
              doubledecker plot\\
              & \codefun{labeling\_lboxed} & left-aligned labels with boxes.\\
              & \codefun{labeling\_left} & left-aligned border labels\\
              & \codefun{labeling\_left2} & left-aligned border
              labels, all labels on top and left border\\
              & \codefun{labeling\_list} & draws a list of labels
              under the plot\\\hline\hline
     shading  & \codefun{shading\_binary} & visualizes the sign of the  residuals\\
              & \codefun{shading\_Friendly} & implements Friendly-like
              shading (based on HSV-colors)\\
              & \codefun{shading\_hcl} & shading based on HCL-colors\\
              & \codefun{shading\_hsv} & shading based on HSV-colors\\
              & \codefun{shading\_max} & shading visualizing the
              maximum test statistic\\ && (based on HCL-colors)\\\hline\hline
     spacing  & \codefun{spacing\_conditional} & increasing spacing
              for conditioning variables,\\&& equal spacing for
              conditioned variables.\\
              & \codefun{spacing\_dimequal} & equal spacing for each dimension\\
              & \codefun{spacing\_equal} & equal spacing for all dimensions\\
              & \codefun{spacing\_highlighting} & increasing spacing,
              last dimension set to zero\\
              & \codefun{spacing\_increase} & increasing spacing\\\hline\hline
     legend   & \codefun{legend\_fixed} & creates a fixed number of
              bins (similar to \codefun{mosaicplot})\\
              & \codefun{legend\_resbased} & suitable for an
              arbitrary number of bins\\&& (also for continuous shadings)\\\hline

  \end{tabular}
  \caption{Available grapcon generators in the strucplot framework}
  \label{tab:grapcons}
\end{table}
